\documentclass[smallabstract,smallcaptions]{dccpaper}

\usepackage[utf8]{inputenc}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{textgreek}
\usepackage{url}

\newlength{\figurewidth}
\newlength{\smallfigurewidth}
\setlength{\smallfigurewidth}{2.75in}
\setlength{\figurewidth}{6in}


% Mathematics
\newcommand{\set}[1]{\ensuremath{\{ #1 \}}}
\newcommand{\abs}[1]{\ensuremath{\lvert #1 \rvert}}
\newcommand{\Oh}{\ensuremath{\mathsf{O}}}
\newcommand{\oh}{\ensuremath{\mathsf{o}}}
\newcommand{\Th}{\ensuremath{\mathsf{\Theta}}}

% Structures
\newcommand{\SA}{\textsf{SA}}
\newcommand{\ISA}{\textsf{ISA}}
\newcommand{\BWT}{\textsf{BWT}}
\newcommand{\FMI}{\textsf{FMI}}
\newcommand{\C}{\textsf{C}}
\newcommand{\RA}{\textsf{RA}}
\newcommand{\mSA}{\ensuremath{\mathsf{SA}}}
\newcommand{\mISA}{\ensuremath{\mathsf{ISA}}}
\newcommand{\mBWT}{\ensuremath{\mathsf{BWT}}}
\newcommand{\mC}{\ensuremath{\mathsf{C}}}
\newcommand{\mRA}{\ensuremath{\mathsf{RA}}}

% Operations
\newcommand{\LF}{\textsf{LF}}
\newcommand{\rank}{\textsf{rank}}
\newcommand{\select}{\textsf{select}}
\newcommand{\mLF}{\ensuremath{\mathsf{LF}}}
\newcommand{\mrank}{\ensuremath{\mathsf{rank}}}
\newcommand{\mselect}{\ensuremath{\mathsf{select}}}

% Shorthands
\newcommand{\Acoll}{\ensuremath{\mathcal{A}}}
\newcommand{\Bcoll}{\ensuremath{\mathcal{B}}}

% Names
\newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
\newcommand{\sdarray}{\textsf{sdarray}}



\begin{document}

\title
{\large
\textbf{Burrows-Wheeler transform for terabases}
}


\author{%
Jouni Sirén \\[0.5em]
{\small\begin{minipage}{\linewidth}\begin{center}
\begin{tabular}{c}
Wellcome Trust Sanger Institute \\
Wellcome Trust Genome Campus \\
Hinxton, Cambridge, CB10 1SA, UK \\
\url{jouni.siren@iki.fi}
\end{tabular}
\end{center}\end{minipage}}
}


\maketitle
\thispagestyle{empty}


\begin{abstract}
Fooo
\end{abstract}


\Section{Introduction}

Sequence data

BWT, compression, FM-index, applications, repetitive, rle

ReadServer; multiple BWTs due to construction issues; less BWTs would mean less memory, less querying

Large scale construction: speed, space, hardware, efficiency

BWT construction: from SA \cite{Mori2008,Nong2011}
external SA \cite{Gonnet1992,Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}
space efficient algorithms \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}
external BWT \cite{Ferragina2012,Beller2013}

Specialized for DNA: \cite{Bauer2013,Liu2014,Li2014a,Pantaleoni2014,Wang2015}

This paper: terabases


\Section{Background}

A \emph{string} $S[1,n] = s_{1} \dotsm s_{n}$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{1, \dotsc, \sigma}$. For indexing purposes, we consider \emph{text} strings $T[1,n]$ terminated by an endmarker $T[n] = \$ = 0$ not occurring anywhere else in the text. \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$. A \emph{substring} of string $S$ is a sequence of the form $S[i,j] = s_{i} \dotsm s_{j}$. Substrings of the type $S[1,j]$ and $S[i,n]$ are called \emph{prefixes} and \emph{suffixes}, respectively.

The \emph{suffix array} (\SA) \cite{Manber1993} is a simple full-text index of a string. Given a text $T$, its suffix array $\mSA_{T}[1,n]$ is an array of pointers to the suffixes of the text in \emph{lexicographic order}. (If the text is evident from the context, we will omit it and write just \SA.) We can build the suffix array in $\Oh(n)$ time using approximately $n(\log n + \log \sigma + 2)$ bits of space, or just $2n$ bits of working space over the text and the suffix array \cite{Nong2011}. Given a \emph{pattern} $P$, we can find the \emph{lexicographic range} $[sp,ep]$ of suffixes of the text prefixed by the pattern in $\Oh(\abs{P} \log n)$ time by binary searching in the suffix array. The corresponding range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.

While the suffix array is simple, it requires several times more memory than the original text. For large texts, this can be a serious drawback. The more space-efficient alternatives to the suffix array are based on the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, an easily reversible permutation of the text with a similar combinatorial structure to the suffix array. Given a text $T[1,n]$ and its suffix array, we can easily produce the \BWT{} as $\mBWT[i] = T[\mSA[i]-1]$ (with $\mBWT[i] = T[n]$, if $\mSA[i] = 1$).

A key observation is that if suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$. Therefore, if $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT, and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order. This forms the essence of \LF\emph{\nobreakdash-mapping}. The suffix preceding $T[\mSA[i],n]$ in text order is the one starting at $\mSA[\mLF(i)]$, where
$$
\mLF(i) = \mC[\mBWT[i]] + \mBWT.\mrank(i, \mBWT[i]),
$$
$\mC[c]$ is the number of suffixes starting with a character smaller than $c$, and $S.\mrank(i,c)$ is the number of occurrences of character $c$ in the prefix $S[1,i]$. For searching, we use a more general definition of \LF\nobreakdash-mapping:
$$
\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
$$
which tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$ in lexicographic order. This is known as the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of string $cT[\mSA[i],n]$ among the suffixes of text $T$.

The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures. The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found through \emph{backward searching}. Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern. We can find $[sp_{i-1},ep_{i-1}]$ as
$$
[sp_{i-1},ep_{i-1}] = [\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])].
$$
By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT. In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding.

In order to find the occurrences of the pattern in the text, the FM-index \emph{samples} some suffix array pointers, and uses \LF\nobreakdash-mapping to derive the unsampled pointers. The set of sampled pointers always includes a pointer to the beginning of each text. If $\mSA[i]$ is not sampled, the FM-index proceeds to $\mLF(i)$ and continues from there. If $\mSA[\mLF^{k}(i)]$ is the first sample encountered, $\mSA[i] = \mSA[\mLF^{k}(i)] + k$. Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.

Assume that we have an ordered \emph{collection} of texts $\Acoll = (T_{1}, \dotsc, T_{m})$ of total length $n = \abs{\Acoll} = \sum_{i} \abs{T_{i}}$. We want to build a (generalized) \BWT{} for the collection. The usual way is to assume that the endmarkers of all text are distinct, with the one at the end of text $T_{i}$ having character value $(0,i)$. This guarantees that each suffix has a unique lexicographic rank among the suffixes of the collection. We still encode each endmarker as $0$ in the \BWT{} in order to save space. Because of this, \LF\nobreakdash-mapping does not work with $c = 0$, and we cannot match patterns spanning text boundaries.

With large collections of short texts (such as short reads), there are more space-efficient alternatives to sampling. Because all endmarkers had distinct character values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the endmarker of text $T_{i}$. To reach the endmarker, we iterate
$$
\Psi(i) = \mBWT.\mselect(i - \mC[c], c),
$$
where $c$ is the largest character value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$. If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as
$$
T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)] \,\, \textrm{for} \,\, 1 \le j < \abs{T_{i}}.
$$
As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also extract the lexicographic ranks of all suffixes of text $T_{i}$ in the process.


\Section{Space-efficient \BWT{} construction}

The FM-index was introduced as a more space-efficient alternative to the suffix array. If we have to build a suffix array first in order to construct an FM-index, we lose a large part of the benefits, and index construction becomes the overall bottleneck. To overcome this bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly. Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text colletions that are too large to fit in memory in an uncompressed form.

Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character. The transformation \cite{Hon2007} is quite straightforward. We first find the pointer $\mSA[i]$ to the beginning of text $T$ (the position $i$, where $\mBWT[i] = 0$). Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$. Finally we transform the \BWT{} by \emph{replacing} the endmarker at $\mBWT[i]$ with the inserted character $c$, and by \emph{inserting} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.

There are six basic ways of using the transformation in a \BWT{} construction algorithm. Instead of inserting a single character at a time, we may use \emph{batch updates} and transform the \BWT{} of text $T$ directly into that of text $XT$, where $X$ is any string \cite{Hon2007}. We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}. We can also have the \BWT{} of a text collection, and \emph{extend} multiple texts by inserting a new character to the beginning of each of them in a single update \cite{Bauer2013}. In all three cases, we may use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT.


\Section{Merging \BWT{}s of text collections}

% FIXME: a running example with 2 texts
Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them. Assume that we store the \BWT{}s in a two-level array, where the first level contains pointers to \emph{blocks} of size $b$ (bits). If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead. With $b = \sqrt{x \log x}$, the overhead becomes $\Oh(\sqrt{x \log x})$ bits.

The algorithm for merging $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ into $\mBWT_{\Acoll \cup \Bcoll}$ has three phases \cite{Siren2009}:
\begin{enumerate}

\item \textbf{Search.} We search for all texts in collection $\Bcoll$ using $\mBWT_{\Acoll}$ and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$. This takes $\Oh(n_{\Bcoll} t_{r})$ time. If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.

\item \textbf{Sort.} We sort the ranks produced in the search phase to build the \emph{rank array} (\RA) of collection $\Bcoll$ relative to collection $\Acoll$. The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $X$ is the suffix pointed by $\mSA_{\Bcoll}[i]$. The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$. If we extract the texts from $\mBWT_{\Bcoll}$ in the search phase, we can write the ranks directly into the rank array, making this phase unnecessary.

\item \textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array. If $\mRA_{\Bcoll \mid \Acoll}[i] = j$, the merged \BWT{} will have $j$ characters from $\mBWT_{\Acoll}$ before $\mBWT_{\Bcoll}[i]$. This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time. By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$ once we have processed them, we can merge the \BWT{}s almost in-place, using $\Oh(\sqrt{x \log x})$ bits of working space, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.

\end{enumerate}
The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$. The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.

Alternatively, we can build the \emph{interleaving bitvector} $B_{\Acoll \cup \Bcoll}$. This bitvector is a binary sequence of length $n_{\Acoll} + n_{\Bcoll}$ that encodes the same information as the rank array. We set $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$, denoting that $\mSA_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]]$ points to a suffix of collection $\Bcoll$. This approach uses $n_{\Acoll} + n_{\Bcoll} + \Oh(\sqrt{x \log x})$ bits of working space.


\Section{Large-scale \BWT{} merging}

If we split text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ by merging the \BWT{}s of the subcollections incrementally. Merging the subcollections takes $\Oh((p+t_{r})n)$ time and uses (essentially) $\min(n, \frac{n}{p} \log n)$ bits of working space.

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection. Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.

The key to reducing the space overhead is that we do not need to keep the rank array in memory. We can write the lexicographic ranks to \emph{disk} in the search phase, sort them to build the rank array in the sort phase, and scan the rank array once during the merge phase. In practice, we may want to \emph{interleave} the sort phase with the search/merge phases and store the lexicographic ranks on disk in a \emph{compressed form}, because the amount of fast disk space can also become a bottleneck. In the following, we describe the key ideas that make large-scale \BWT{} merging fast and space-efficient.

\SubSection{Search}

Instead of searching for every text in collection $\Bcoll$ separately, we can search for the \emph{reverse trie} of the collection. Assume that there are $m_{\Acoll}$ texts in collection $\Acoll$ and $m_{\Bcoll}$ texts in collection $\Bcoll$. The \emph{root} of the trie corresponds to suffix $\$$, which has lexicographic rank $m_{\Acoll}$ in $\Acoll$ and corresponds to lexicographic range $[1,m_{\Bcoll}]$ in $\Bcoll$.

% FIXME pseudocode if we have space for it
Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$. This means that suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, so we can output a \emph{run} of lexicographic ranks $(r, sp+1-ep)$. Afterwards, we proceed to the \emph{children} of the node. For each character $c \in \Sigma$, we create a node corresponding to suffix $cX$, lexicographic rank $\mLF_{\Acoll}(r,c)$, and lexicographic range $[\mLF_{\Bcoll}(sp-1, c) + 1, \mLF_{\Bcoll}(ep, c)]$. Searching the branches of the trie can be done in parallel using multiple \emph{threads}.

\SubSection{Buffering}

To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk. Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}. The run buffer stores the runs as pairs of integers $(r, \ell)$. Once the run buffer becomes full, we sort the runs by \emph{run heads} $r$, use \emph{differential encoding} for the run heads, and encode the differences and run lengths with a \emph{prefix-free code}. The compressed run buffer is then merged with the similarly compressed thread buffer.

Once a thread buffer becomes full, we insert it into the global \emph{merge buffers}. There are $k$ merge buffers, with buffer $i$ containing $2^{i-1}$ merged thread buffers. The insertion starts at buffer $1$. If buffer $i$ is empty, the thread inserts its thread buffer there and returns back to the search phase. Otherwise it takes the contents of merge buffer $i$, merges them with its thread buffer, and proceeds to buffer $i+1$. If a thread reaches buffer $k+1$, it writes its thread buffer to disk and returns back to work.

\SubSection{Merge}

The merge phase starts with the rank array stored in multiple files on disk. Each of the files is sorted and compressed. For interleaving the \BWT{}s, we need to merge the files to produce an increasing run-length encoded stream of lexicographic ranks.

This work can also be parallelized between multiple threads. One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks. Another thread consumes the stream and uses it to interleave the \BWT{}s. Because the producer thread is usually slower than the consumer thread, we may want to use multiple threads for the multiway merge, if the disk is fast enough.


\Section{Implementation}

We have implemented the proposed enhancements to the \BWT{} merging algorithm in a tool (\BWTmerge) intended for merging the \BWT{}s of large collections of short reads. \BWTmerge{} is written in C++, and the source code is available at GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation is built on top of the \emph{SDSL library} \cite{Gog2014b} and uses the features of C++11 extensively. As a result, it needs a fairly recent C++11 compiler to compile. We have built \BWTmerge{} on Linux and OS~X using g++.

The target environment of \BWTmerge{} is a \emph{single node} of a \emph{computer cluster}. The system should have tens of CPU cores and hundreds of gigabytes of memory. The amount of local disk space might not be much larger than memory size, while there can be plenty of shared disk space available. The number of search threads is equal to the number of CPU cores, while the merge phase uses just one producer thread and one consumer thread. By adjusting the sizes of run buffers and thread buffers and the number of merge buffers, \BWTmerge{} should work reasonably well in different environments.

The internal alphabet of \BWTmerge{} is \texttt{012345}, which corresponds to either \texttt{\$ACGTN} or \texttt{\$ACGNT}, depending on where the \BWT{}s come from. \BWT{}s using different alphabetic orders cannot be merged. We use simple byte-level codes for \emph{run-length encoding} the \BWT{}s. The encoding of run $(c, \ell)$, where $c$ is the character value and $\ell$ is the length, depends on the length of the run. If $\ell \le 41$, the run is encoded in a single byte as $6 \cdot (\ell-1) + c$. Longer runs start with byte $6 \cdot 41 + c$, followed by the encoding of $\ell-42$. The remaining run length is encoded as a sequence of bytes, with the low 7 bits containing data and the high bit telling whether the encoding continues in the next byte. The compressed buffers use the same 7+1\nobreakdash-bit code for both differentially encoded run heads and run lengths.

For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, so that no run continues from one block to the next one. For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$. Each of these seven increasing sequences is encoded using the \sdarray{} encoding \cite{Okanohara2007}. To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block. A \select{} query on the same sequence transforms position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block. Finally, we decompress the block to answer the query. \select{} queries and random accesses to the \BWT{} work in a similar way. There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a relatively short lexicographic range.

We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}. In our experiments, this direct memory management reduced the space overhead by tens of gigabytes. When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB. Each thread tries to reuse the heap it used for the last allocation. If a heap is full or the thread cannot acquire the mutex, it moves to the next heap. With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle. Most calls to \texttt{free()} could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.


\Section{Experiments}

environment, compiler

data: ReadServer BWTs, NA12878 trio \cite{1000GP2015}
error correction using bfc \cite{Li2015}

ReadServer: trade-offs with different parameters

trio: ropebwt + this vs ropebwt vs ropebwt2


\Section{Discussion}

conclusions

sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information

remove duplicates


\Section{References}
\bibliographystyle{IEEEtran}
\bibliography{paper}


\end{document}
